Gastro-intestinal transmission and colonisation of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model

### 2025-10-09 paper/paper.qmd
hid non stochastic figures

```{r fig-sensitivity-analysis, include=TRUE, eval=TRUE}
#| echo: false
#| fig-cap: "Sensitivity analysis of the model to changes in key infection control parameters. The y axis shows the total number of nosocomial infections during the outbreak period. Antimicrobial prescription rate represents the percentage of patients that were administered an antimicrobial during their admission. Proportion redistributed represents the proportion of infected patients that were moved to a single-occupancy room (lower values represent fewer patients being moved). Toilet cleaning effect represents the effectiveness of cleaning in reducing toilet contamination (100% = complete removal of contamination; 0% = no cleaning effect). Toilet cleaning rate represents the number of times toilets were cleaned per day, on average."
list(
  abx_prescription_rate    = sens_abx_prescription_outbreak_cases,
  proportion_redistributed = sens_prop_redist_outbreak_cases,
  toilet_cleaning_rate     = sens_toilet_cleaning_rate_outbreak_cases,
  toilet_cleaning_effect   = sens_toilet_cleaning_effect_outbreak_cases
) %>%
  bind_rows() %>% 
  pivot_longer(
    cols = starts_with("c-"),
    names_to = "param",
    values_drop_na = TRUE
  ) %>% 
  group_by(param, value) %>% 
  summarise(
    mean_cases = mean(`outbreak-cases`),
    sd_cases   = sd(`outbreak-cases`),
    .groups    = "keep"
  ) %>% 
  mutate(
    # convert antibiotic prescription rate to 0-100 scale, leave others unchanged
    # value = if_else(param == "c-antibiotic-prescription-rate", value * 100, value),
    param = case_when(
      param == "c-antibiotic-prescription-rate" ~ "Antimicrobial prescription rate",
      param == "c-side-room-delay-mean"     ~ "Side room delay mean",
      param == "c-toilet-cleaning-rate"         ~ "Toilet cleaning rate",
      param == "c-toilet-cleaning-effect"       ~ "Toilet cleaning effect",
      TRUE                                      ~ param
    )
  ) %>% 
  ggplot(aes(x = value, y = mean_cases)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = mean_cases - sd_cases, ymax = mean_cases + sd_cases), alpha = 0.2) +
  facet_wrap(~ param, scales = "free") +    # free_x required to use facetted_pos_scales
  # apply percentage labels and a per-facet x-title only for the antibiotic facet
  ggh4x::facetted_pos_scales(
    x = list(
      # LHS is evaluated against the facet layout; RHS is the scale to use for matching facets
      param %in% c("Antimicrobial prescription rate", "Proportion redistributed", "Toilet cleaning effect") ~
        scale_x_continuous(
          labels = scales::label_percent(scale = 100)   # data already on 0-100 so use scale=1
        )
    )
  ) +
  # fallback labels (these apply to facets that did not get a custom scale)
  labs(x = "Parameter value",
       y = "Number of infected patients")
```

@@

```{r run-deterministic-model, include=FALSE, cache=TRUE}
# Parameters and time-dependent functions
parameters <- list(
  # p = 1 - unique(mean_sims$`community-colonisation-rate`),      # Proportion of non-infected admissions
  A = 50        # Total admission/discharge rate
)
#

@@

```{r run-stochastic-sd-model, include=FALSE, cache=TRUE}
set.seed(42)
randomLHS_sd <- randomLHS(1000000, 8)
sd_param_grid <- data.frame(init_population = rep(610, 1000),
                            A = 68,
                            r = qunif(randomLHS_sd[, 3], 0.001, 0.01),
                            baseline_beta = qunif(randomLHS_sd[, 4], 0.0001, 0.01),
                            baseline_p = qunif(randomLHS_sd[, 5], 0.001, 0.1),
                            outbreak_beta = qunif(randomLHS_sd[, 6], 0.0001, 0.01),
                            outbreak_p = qunif(randomLHS_sd[, 7], 0.001, 0.1),
                            outbreak_start = rep(1042, 1000),
                            outbreak_end = rep(1234, 1000),
                            n_days = rep(2585, 1000))

@@

# abc
abc_sd <- abc(salgado$rates,
              sd_param_grid, dplyr::select(sd_sumstats, -param_row), tol=0.0025, method='rejection')

@@

posterior_predictive_sim_sd <- 
  pmap_dfr(posterior_check_params,
        \(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days, param_row, seed) {
            set.seed(seed)
            sd_stochastic(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days) %>% 
              mutate(`random-seed` = seed,
                     param_row = param_row) %>% 
              bind_rows()
        }, .progress = T)

@@

```{r fig-single-outbreak-deterministic-one-plot, include=FALSE}
#| echo: false
#| message: false
#| fig.cap: "Comparison of infection rates produced by the agent-based model and the deterministic model. The r"
single_outbreak_one_seed %>% 
  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
  filter(`[step]` > 31) %>%
  group_by(compartment, `[step]`) %>% 
  summarise(m = mean(value)) %>% 
  ggplot(aes(x = `[step]`, y = m, color = compartment)) +
  geom_line() +
  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
       x = "Time",
       y = "Number of Patients",
       color = "Compartment") +
  geom_line(data = output_df, aes(x = time, y = N), color = "blue", linetype = "dashed") + 
  geom_line(data = output_df, aes(x = time, y = I), color = "red", linetype = "dashed") +
  xlim(500, 2000)
```

```{r fig-single-outbreak-deterministic-faceted, include=TRUE}
#| echo: false
#| message: false
#| fig.cap: "Comparison of infection rates produced by the agent-based model and the deterministic model. The parameters for the deterministic model were determined using approximate Bayesian computation, using the output from a single run of the agent-based model as a reference."
output_df_long <- output_df %>% 
  pivot_longer(cols = c(N, I), names_to = "compartment", values_to = "value") %>% 
  mutate(sim_type = "Deterministic") %>% 
  mutate(compartment = case_when(
    compartment == "N" ~ "Not infected",
    compartment == "I" ~ "Infected"
  ))

@@

```{r fig-abc-sd-posteriors, include=TRUE}
#| echo: false
#| fig.cap: "Posterior distributions of parameters obtained from approximate Bayesian computation using summary statistics generated from the stochastic deterministic model. The red line shows the prior distribution; the blue line shows the posterior distribution."
```

>>>>

```{r fig-sensitivity-analysis, include=TRUE, eval=TRUE}
#| echo: false
#| fig-cap: "Sensitivity analysis of the model to changes in key infection control parameters. The y axis shows the total number of nosocomial infections during the outbreak period. Antimicrobial prescription rate represents the percentage of patients that were administered an antimicrobial during their admission. Side room delay is the mean number of days an infected patient remained in a bay before being moved to a side room. Toilet cleaning effect represents the effectiveness of cleaning in reducing toilet contamination (100% = complete removal of contamination; 0% = no cleaning effect). Toilet cleaning rate represents the number of times toilets were cleaned per day, on average."
list(
  abx_prescription_rate    = sens_abx_prescription_outbreak_cases,
  proportion_redistributed = sens_prop_redist_outbreak_cases,
  toilet_cleaning_rate     = sens_toilet_cleaning_rate_outbreak_cases,
  toilet_cleaning_effect   = sens_toilet_cleaning_effect_outbreak_cases
) %>%
  bind_rows() %>% 
  pivot_longer(
    cols = starts_with("c-"),
    names_to = "param",
    values_drop_na = TRUE
  ) %>% 
  group_by(param, value) %>% 
  summarise(
    mean_cases = mean(`outbreak-cases`),
    sd_cases   = sd(`outbreak-cases`),
    .groups    = "keep"
  ) %>% 
  mutate(
    # convert antibiotic prescription rate to 0-100 scale, leave others unchanged
    # value = if_else(param == "c-antibiotic-prescription-rate", value * 100, value),
    param = case_when(
      param == "c-antibiotic-prescription-rate" ~ "Antimicrobial prescription rate",
      param == "c-side-room-delay-mean"     ~ "Side room delay mean (days)",
      param == "c-toilet-cleaning-rate"         ~ "Toilet cleaning rate (per day)",
      param == "c-toilet-cleaning-effect"       ~ "Toilet cleaning effect",
      TRUE                                      ~ param
    )
  ) %>% 
  ggplot(aes(x = value, y = mean_cases)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = mean_cases - sd_cases, ymax = mean_cases + sd_cases), alpha = 0.2) +
  facet_wrap(~ param, scales = "free") +    # free_x required to use facetted_pos_scales
  # apply percentage labels and a per-facet x-title only for the antibiotic facet
  ggh4x::facetted_pos_scales(
    x = list(
      # LHS is evaluated against the facet layout; RHS is the scale to use for matching facets
      param %in% c("Antimicrobial prescription rate", "Proportion redistributed", "Toilet cleaning effect") ~
        scale_x_continuous(
          labels = scales::label_percent(scale = 100)   # data already on 0-100 so use scale=1
        )
    )
  ) +
  # fallback labels (these apply to facets that did not get a custom scale)
  labs(x = "Parameter value",
       y = "Number of infected patients")
```

@@

```{r run-deterministic-model, eval=FALSE, include=FALSE, cache=TRUE}
# Parameters and time-dependent functions
parameters <- list(
  # p = 1 - unique(mean_sims$`community-colonisation-rate`),      # Proportion of non-infected admissions
  A = 50        # Total admission/discharge rate
)
#

@@

```{r run-stochastic-sd-model, include=FALSE, cache=TRUE, cache.lazy=FALSE}
set.seed(42)
randomLHS_sd <- randomLHS(1000000, 8)
sd_param_grid <- data.frame(init_population = rep(610, 1000),
                            A = 68,
                            r = qunif(randomLHS_sd[, 3], 0.001, 0.01),
                            baseline_beta = qunif(randomLHS_sd[, 4], 0.0001, 0.01),
                            baseline_p = qunif(randomLHS_sd[, 5], 0.001, 0.1),
                            outbreak_beta = qunif(randomLHS_sd[, 6], 0.0001, 0.01),
                            outbreak_p = qunif(randomLHS_sd[, 7], 0.001, 0.1),
                            outbreak_start = rep(1042, 1000),
                            outbreak_end = rep(1234, 1000),
                            n_days = rep(2585, 1000))

@@

# abc
abc_sd <- abc(salgado$rates,
              dplyr::select(sd_param_grid, -param_row, n_days),
              dplyr::select(sd_sumstats, -param_row), tol=0.0025, method='rejection')

@@

posterior_predictive_sim_sd <- 
  pmap_dfr(posterior_check_params,
        \(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days, seed) {
            set.seed(seed)
            sd_stochastic(init_population, A, r, baseline_beta, baseline_p, outbreak_beta, outbreak_p, outbreak_start, outbreak_end, n_days) %>% 
              mutate(`random-seed` = seed) %>% 
              bind_rows()
        }, .progress = T)

@@

```{r fig-single-outbreak-deterministic-one-plot, include=FALSE, eval=FALSE}
#| echo: false
#| message: false
#| fig.cap: "Comparison of infection rates produced by the agent-based model and the deterministic model. The r"
single_outbreak_one_seed %>% 
  pivot_longer(cols = c(`current-colonised`, `current_not_infected`), names_to = "compartment", values_to = "value") %>%
  filter(`[step]` > 31) %>%
  group_by(compartment, `[step]`) %>% 
  summarise(m = mean(value)) %>% 
  ggplot(aes(x = `[step]`, y = m, color = compartment)) +
  geom_line() +
  labs(title = "Dynamics of Non-Infected and Infected Patients (Time-Varying Beta)",
       x = "Time",
       y = "Number of Patients",
       color = "Compartment") +
  geom_line(data = output_df, aes(x = time, y = N), color = "blue", linetype = "dashed") + 
  geom_line(data = output_df, aes(x = time, y = I), color = "red", linetype = "dashed") +
  xlim(500, 2000)
```

```{r fig-single-outbreak-deterministic-faceted, include=FALSE, eval=FALSE}
#| echo: false
#| message: false
#| fig.cap: "Comparison of infection rates produced by the agent-based model and the deterministic model. The parameters for the deterministic model were determined using approximate Bayesian computation, using the output from a single run of the agent-based model as a reference."
output_df_long <- output_df %>% 
  pivot_longer(cols = c(N, I), names_to = "compartment", values_to = "value") %>% 
  mutate(sim_type = "Deterministic") %>% 
  mutate(compartment = case_when(
    compartment == "N" ~ "Not infected",
    compartment == "I" ~ "Infected"
  ))

@@

```{r fig-abc-sd-posteriors, include=TRUE}
#| echo: false
#| fig.cap: "Posterior distributions of parameters obtained from approximate Bayesian computation using summary statistics generated from the stochastic deterministic model. The red line shows the prior distribution; the blue line shows the posterior distribution."

sd_posterior_params_abc <- abc_sd$unadj.values
colnames(sd_posterior_params_abc) <- names(dplyr::select(sd_param_grid, -param_row))
sd_posterior_params_abc %>% 
  as.data.frame() %>% 
  dplyr::select(r, baseline_beta, baseline_p, outbreak_beta, outbreak_p) %>% 
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") %>% 
  mutate(parameter = factor(parameter, 
                            levels = c("r", "baseline_beta", "outbreak_beta", "baseline_p", "outbreak_p"))) %>% 
  ggplot(aes(x = value)) +
  geom_density(fill = "blue", alpha = 0.5) +
  facet_wrap(~ parameter, scales = "free")
